Agile Monitoring of Evolving Web Services: For the Little Man

### 2018-04-06 evolution.tex
Edits to Sections 2 to 6

  \section{Version-Aware Monitoring}

@@

  Version control in \tool can be supported in two ways,
  the developer explicitly states the current version, 
  or the current version is automatically detected based
  on some version control system. 
  The second option is the most interesting one.
  For the systems deployed using Git, an extra configuration line enables \tool to automatically\footnote{Alternatively, the maintainer can add version identifiers manually for the web application through a configuration file if the system does not use git.} associate API traffic with the currently deployed version based on information in the .git folder \footnote{\url{https://git-scm.com/}}:

@@

    \begin{figure}[h!]
      \centering
      \includegraphics[width=0.9\linewidth]{utilization-evolution}
      \caption{The Evolution of All the API Endpoints Utilization Across System Versions}
      \label{fig:mv-util}
    \end{figure}

@@

    \item \epFeedItems is discontinued a little bit after \epTopArticles is being introduced

@@

\subsection*{Evolving Grouped Performance}
  The limitation of the previous view is that it does not present the information also on a per version basis. To address this, a different visual perspective entitled \perspective{User-Focused Multi-Version Endpoint Performance} can be defined. Figure \ref{fig:tuv} presents such a perspective by mapping the average execution time for a given user (lines) and given version (columns) on the area of the corresponding circle.

@@

The colors represent users. The figure shows average performance varying  across users and versions with no clear trend: this is probably because varying user workload (i.e. number of sources to which the user is registered) is the reason for the variation in response times. \ins{one can see that for user 1 performance degrades over the versions. Given that for other users the performance does not degrade in the same way, it is probable that the problem might lay elsewhere: the server was overloaded or the endpoint is ``algorithmically slower''.}

>>>>

  \section{Version-Aware Monitoring}
  \label{sec:version}

@@

  Version control in \tool can be supported in two ways:
  the developer explicitly states the current version, 
  or the current version is automatically detected based
  on some version control system. 
  The second option is the most interesting one.
  For the systems deployed using Git, an extra configuration line enables \tool to automatically\footnote{Alternatively, the maintainer can add version identifiers manually for the web application through a configuration file if the system does not use git.} associate API traffic with the currently deployed version based on information in the .git folder\footnote{\url{https://git-scm.com/}}:

@@

    \begin{figure}
      \centering
      \includegraphics[width=0.9\linewidth]{utilization-evolution}
      \caption{The Evolution of All the API Endpoints Utilization Across System Versions}
      \label{fig:mv-util}
    \end{figure}

@@

    \item \epFeedItems is discontinued a little bit after \epTopArticles is being introduced
	\item TBA

@@

\subsection*{Evolving Grouped Performance}
  The limitation of the previous view is that it does not present the information also on a per user basis. To address this, a different visual perspective entitled \perspective{User-Focused Multi-Version Endpoint Performance} can be defined. \Fref{fig:tuv} presents such a perspective by mapping the average execution time for a given user (lines) and given version (columns) on the area of the corresponding circle.

@@

The colors represent users. The figure shows average performance varying  across users and versions with no clear trend: this is probably because varying user workload (i.e. number of sources to which the user is registered) is the reason for the variation in response times. \ins{One can see that for user with id=1, performance degrades over consecutive versions. Given that for other users the performance does not degrade in the same way, it is probable that the problem might lay elsewhere: the server was overloaded or the endpoint is ``algorithmically slower''.} \va{could it be that the user kept on adding sources to their profile?}

### 2018-04-06 grouping.tex
Edits to Sections 2 to 6

\section{Grouping Requests}
\label{sec:user}

@@

\begin{itemize}
	\item \code{'User'} stands for the name of this particulart grouping strategy
	\item \code{lambda: Session.find(... } is a callback with no arguments, 
	that makes use of the global flask.request and extracts from 
	the request the current session, and in turn the user id\footnote{
		Note that every web service or application must have a way of 
		associating a request with a user. In fact, the Session.find(request) 
		was already an existing function in the analyzed system}
\end{itemize}

@@

In Zeeguu, \epFeedItems retrieves a list of recommended articles for a given user. Cf. \Fref{fig:ep} it is the endpoint with the slowest response time and highest variability. The reason for this is that a user can be subscribed to anything from one to three dozen article sources and for each of the sources the system must compute the personalized difficulty of each article at every request.

\begin{figure}[h!]
  \centering
  \includegraphics[width=.5\linewidth]{time_per_user}
  \caption{The \epFeedItems shows a very high variability across users}
  \label{fig:tpu}
\end{figure}

A \perspective{Per-User Performance} perspective should show the different response times for different users. Figure \ref{fig:tpu} presents a subset of the corresponding view in the \tool. The figure shows that the response times for this endpoint can vary considerably for different users with some extreme cases where a user has to wait a full minute until their recommended articles are shown\footnote{After seeing this perspective, the maintainer refactored the architecture of the system to move part the difficulty computation out of the interactive loop}.

>>>>

\section{Per-User Performance Monitoring}
\label{sec:user}

@@

\begin{itemize}
	\item \code{'User'} stands for the name of this particular grouping strategy
	\item \code{lambda: Session.find... } is a callback with no arguments, 
	that makes use of the global \code{flask.request} and extracts from 
	the request the current session, and in turn the user id\footnote{
		Note that every web service or application must have a way of 
		associating a request with a user. In fact, the Session.find(request) 
		was already an existing function in the analyzed system}
\end{itemize}

@@

In Zeeguu, the \epFeedItems endpoint retrieves a list of recommended articles for a given user. As shown in \Fref{fig:ep}, this is the endpoint with the slowest response time and highest variability. The reason for this is that a user can be subscribed to anything from one to three dozen article sources and for each of the sources the system must compute the personalized difficulty of each article at every request.

\begin{figure}[h!]
  \centering
  \includegraphics[width=.7\linewidth]{time_per_user}
  \caption{The \epFeedItems shows a very high variability across users}
  \label{fig:tpu}
\end{figure}

A \perspective{Per-User Performance} perspective should show the different response times for different users. \Fref{fig:tpu} presents a subset of the corresponding view in the \tool. The figure shows that the response times for this endpoint can vary considerably for different users with some extreme cases where a user has to wait a full minute until their recommended articles are shown\footnote{After seeing this perspective, the maintainer refactored the architecture of the system to move part the difficulty computation out of the interactive loop}.

### 2018-04-06 introductory.tex
...
Edits to Sections 2 to 6

\section{Case Study: The API (Zeeguu)}
\label{sec:case}

%\todo{Either leave as is or add more info on usage by college if possible}

@@

    \item Reader applications that provide effortless translations for those texts which are too difficult for the readers.

    \item Interactive exercises personally generated based on the preferences and past likes of the learner.

    \item Article recommendations which are personalized for the interests of the learner and come with difficulty estimation that helps the learner find articles with the appropriate difficulty.

@@

  The core API implemented with Flask and Python provides correspondingly three types of functionality: contextual translations, article recommendations, and personalized exercise suggestions. In total the API provides a bit less than 50 endpoints, out of which probably a dozen are very frequently used. 
  % The development of the core API itself is a research project.

  At the time of writing this article, the ecosystem consists of a reader web application, a web based exercises platform, and a smartwatch application, which are used at the moment of writing this article by more than \activeUserCount active users. The users come from a highschool, a language school, and some users are using it on their own, without any educational context. The highest load we observed until now on the API consisted of 12K requests in one day.

  \ml{add more numbers that show how impressive is our API}

  We will use this \zee API as a case study for this paper. 
  All the figures in this paper are captured from the actual deployment of \tool in the context of the \zee API. The figures are interactive offering basic data exploration capabilities: filter, zoom, and details on demand\cite{Shne99a}. The \tool deployment for the case study can be accessed publicly\footnote{\url{https://zeeguu.unibe.ch/api/dashboard}; username: {\em guest}, password: {\em soap-sac}}.

@@

\section{Case Study: The Technology (Flask)}
\label{sec:tool}

 Flask is a microframework for Python. It IS USED BY THOUSANDS OF PROJECTS. ELABORATE MORE ON THIS... 
 \vspace{2cm}

 Flask depends on two external libraries: the Jinja2 template engine and the Werkzeug WSGI toolkit.

 “Micro” does not mean that your whole web application has to fit into a single Python file (although it certainly can), nor does it mean that Flask is lacking in functionality. The “micro” in microframework means Flask aims to keep the core simple but extensible. Flask won’t make many decisions for you, such as what database to use. Those decisions that it does make, such as what templating engine to use, are easy to change. Everything else is up to you, so that Flask can be everything you need and nothing you don’t.

 Flask has many configuration values, with sensible defaults, and a few conventions when getting started. By convention, templates and static files are stored in subdirectories within the application’s Python source tree, with the names templates and static respectively. While this can be changed, you usually don’t have to, especially when getting started.

 A Python web application based on WSGI has to have one central callable object that implements the actual application. In Flask this is an instance of the Flask class. Each Flask application has to create an instance of this class itself and pass it the name of the module, but why can’t Flask do that itself?

@@

\subsection*{ENSURE CLARITY: NOTHING PREVENTS GENERALIZATION}
In this article we take as an example the Flask case study
and of the \tool as a dashboard. 
But there is nothing that prevents the generalization of 
our approach to other languages and technologies.

>>>>

\section{Case Study}
\label{sec:case}

%\todo{Either leave as is or add more info on usage by college if possible}
\subsection*{Case Study: The API (Zeeguu)}
\label{sec:api}

@@

    \item \textit{Reader applications} that provide effortless translations for those texts which are too difficult for the readers.

    \item \textit{Interactive exercises} that are personally generated based on the preferences and past likes of the learner.

    \item \textit{Article recommendations} which are also personalized for the interests of each learner and come with a difficulty estimation that helps the learner find articles with appropriate difficulty.

@@

  The core API is implemented with Flask and Python provides and correspondingly three types of functionality: contextual translations, personalized exercise suggestions, and article recommendations. In total, the API provides a bit under 50 endpoints, out of which around a dozen are very frequently used. 
  % The development of the core API itself is a research project. 
  %
  At the time of writing, the ecosystem consists of a reader web application, a web-based exercises platform, and a smartwatch application, which are used by more than \activeUserCount active users on a regular basis. 
  %The majority of the users come from a high school and a language school in the city of Groningen. However, some users are using it on their own, outside of any formal educational context. The highest load we observed until now on the API consisted of 12K requests in a single day.

  \ml{add more numbers that show how impressive is our API}
  \va{I don't think this is necessary - I am also moving the user characteristics in the respective subsection below}

  We will use this \zee API as a case study for this paper. 
  All the figures and measurements in this paper are captured from the actual deployment of \tool in the context of the \zee API. The figures are interactive offering basic data exploration capabilities: filter, zoom, and details on demand\cite{Shne99a}. The \tool deployment for the case study can be accessed publicly\footnote{\url{https://zeeguu.unibe.ch/api/dashboard}; username: {\em guest}, password: {\em soap-sac}}.

@@

\subsection*{Case Study: The Technology (Flask)}
\label{sec:flask}

 \va{Consider this subsection moving before the introduction of Zeeguu; it would make a smoother reading}

 Flask is a microframework for Python. It IS USED BY THOUSANDS OF PROJECTS. ELABORATE MORE ON THIS... 
 %\vspace{2cm}
 \va{Actually, you probably have already introduced Flask in the first section, so you just need the first sentence: As discussed in the previous section, Flask is a microframework for Python.}
%
% Flask depends on two external libraries: the Jinja2 template engine and the Werkzeug WSGI toolkit. 
%
 ``Micro'' does not mean that a whole web application has to fit into a single Python file (although it certainly can), nor does it mean that Flask is lacking in functionality. The ``micro'' in microframework means Flask aims to keep the core simple but extensible. Flask won’t make many decisions for the user, such as what database to use. Those decisions that it does make, such as what templating engine to use, are easy to change. Everything else is up to the user.%, so that Flask can be everything you need and nothing you don’t.

 Flask has many configuration values, with sensible defaults, and a few conventions when getting started. By convention, templates and static files are stored in subdirectories within the application's Python source tree, with the names templates and static respectively. While this can be changed, it usually don’t have to, especially when getting started.
%
 A Python web application based on WSGI has to have one central callable object that implements the actual application. In Flask this is an instance of the Flask class. Each Flask application has to create an instance of this class itself and pass it the name of the module, but why can’t Flask do that itself?
 \va{Do you really need the part in the previous sentence after the comma? Just ref the listing}

@@

\subsection*{Study characteristics}

The \tool has been deployed with the case study API for 
ten months, time in which more than 112.000 hits of the API have 
been recorded. \va{which ten months?}
The majority of the users come from a high school and a language school in the city of Groningen using \zee to study foreign languages. However, some users are using it on their own, outside of any formal educational context. The highest load we observed until now on the API consisted of 12K requests in a single day. \va{not sure what can be added here; maybe that's enough}

\subsection*{Role of Authors in the Case Study}
\va{not sure what else you would like to add here...}
One of the authors leads the developer team that develops the \zee API.

Another one of the authors is the main developer in a team 
that develops the \tool.

\subsection*{TO ARGUE: NOTHING PREVENTS GENERALIZATION}
\va{Move this to the discussion section - it's premature to discuss this here} 
In this article we take as an example the Flask case study
and of the \tool as a dashboard. 
But there is nothing that prevents the generalization of 
our approach to other languages and technologies.

### 2018-04-06 loc1.tex
...
Edits to Sections 2 to 6

  The tool is designed to take advantage of the fact that it runs in the same web application context and that the monitored API already has a web presence. It thus simply makes available one extra endpoint (i.e. \code{/dashboard}) at the same address as the API, endpoint which serves the interactive application presented in the remainder of this paper.

  A dashboard can be attached to an already existing Flask application with a single line of code as the code in the following snippet shows
  \footnote{We do not count the import statement that enables that line}:

  % which binds the dashboard to a Flask application\footnote{\ins{ In this paper we present the integration with APIs written in Python and Flask hoping that this will not prevent the reader from seeing the more general idea; all the tools we show here for Flask can be applied to other API technologies (e.g. Django) by simply providing a few back-end adapters in the right places. }}:

  % caption=Configuring the \tool is straightforward,
  \begin{lstlisting}[style=custompython]
  import flask_monitoringdashboard as dashboard

@@

@@

    \begin{figure}[h!]
      \centering
      \includegraphics[width=\linewidth]{selecting_endpoints.png}
      \caption{Once connected to an API the Dashboard presents the endpoints that are available for monitoring}
      \label{fig:sep}
    \end{figure}

  By default the dashboard takes an {\em opt-in} approach to monitoring: to prevent the performance penalties incurred by the dashboard\footnote{We discuss performance issues later} to affect performance sensitive endpoints, the service developer is asked to manually add the endpoints they want to monitor.

@@

    \item {\bf Utilization} -- refers to how the third parties use the API, which parts are most used, which are little used, etc. This is known to be critical information for upstream developer in general\cite{Haen14a}. In the case of source code dependencies one can detect such information by analyzing software repositories. However, in the case of services, there is no other way but monitoring service utilization.

    \item {\bf Performance} as measured in response times. This is important, especially for APIs that are supposed to be integrated in live systems.

>>>>

  The \tool is designed to take advantage of the fact that it runs in the same context as the web application it monitors, and that the monitored API already has a web presence. It thus simply makes available one extra endpoint (i.e. \code{/dashboard}) at the same address as the API, which serves the interactive application presented in the remainder of this paper.

  A dashboard can be attached to an already existing Flask application with a single line of code~\cite{vogel2017low} as the following snippet shows\footnote{We do not count the import statement that enables that line.}:

  \begin{lstlisting}[style=custompython]
  import flask_monitoringdashboard as dashboard

@@

  \va{\Fref{fig:sep} needs to be updated --- it still has 2017 dates}

@@

    \begin{figure}[h]
      \centering
      \includegraphics[width=\linewidth]{selecting_endpoints.png}
      \caption{Once connected to an API the Dashboard presents the endpoints that are available for monitoring}
      \label{fig:sep}
    \end{figure}

  By default the dashboard takes an {\em opt-in} approach to monitoring: to prevent the performance penalties incurred by the dashboard\footnote{We discuss performance issues later.} to affect performance-sensitive endpoints, the service developer is asked to manually define the endpoints they want to monitor.

@@

    \item {\bf Utilization} -- refers to how the third parties use the API, which parts are most used, which are little used, etc. This is known to be critical information for upstream developers in general\cite{Haen14a}. In the case of source code dependencies one can detect such information by analyzing software repositories. However, in the case of services, there is no other way but monitoring service utilization.

    \item {\bf Performance} as measured in response times. Knowing the performance of one's API is critical for the quality of a service API, and especially so for APIs that are supposed to be integrated in live systems.

### 2018-04-06 performance.tex
intro to evolution
Edits to Sections 2 to 6

  \item In the top three slowest endpoints, but the second most used endpoint in the system (cf. \Fref{fig:aeu}) is \epTranslationsColor. The endpoint is critical being embedded in the interactive loop in the reader applications that rely on the Zeeguu API: a user would tap on a word and receive a translation from this endpoint.

@@

\begin{figure}[!ht]
 \centering
 \includegraphics[width=0.85\linewidth]{endpoint_performance_}
 \caption{The response time (in ms) per monitored endpoint view allows for identifying performance variability and balancing issues}
 \label{fig:ep}
\end{figure}

Such a view is important for allocating optimization efforts. In our case study, the API developers started by improving the performance of \epTranslationsColor and then later optimized the performance of \epFeedItemsColor. However, in order to understand the problems of the two endpoints, a big component represents understanding the outliers: because they are many.

  \subsection{Automated Outlier Detection and Monitoring}

  When an API is called from within a highly interactive application (as it is the case with the case study in this paper) 
  of particular interest to the API developers are performance {\em outliers}.

  Indeed, a translation request that takes three times more than expected can seriously decrease the perceived quality of the application. Thus, identifying, collecting all appropriate data, and diagnosing the root causes of such outliers is especially critical in improving the quality of an application.

@@

  \begin{figure}[h!]
    \centering
    \includegraphics[width=0.9\linewidth]{outlier-annotated}
    \caption{Automatically collected outlier information}
    \label{fig:figure1}
  \end{figure}

The bottom panel shows the stack trace. 
In this particular case, it is revealing for the developer to learn that at the time of the stack trace snapshot, the code was in the google\_translator.py: indeed, the system uses as back-end multiple translators, and it has been observed that many of the outliers happen while the system is waiting for the Google translator.

This information has to be corroborated with the observations that neither the memory nor the processor are overloaded at the moment. Thus this functionality in microsoft\_translator is really slow in itself, and this is not a result of the machine being overloaded for example.

>>>>

  \item In the top three slowest endpoints, but the second most used endpoint in the system (cf. \Fref{fig:aeu}) is \epTranslationsColor. The endpoint is critical for the reader applications that rely on the Zeeguu API: each time a user taps on a word on their screen they receive a translation from this endpoint.

@@

\begin{figure}
 \centering
 \includegraphics[width=0.85\linewidth]{endpoint_performance_}
 \caption{The response time (in ms) per monitored endpoint view allows for identifying performance variability and balancing issues}
 \label{fig:ep}
\end{figure}

Such a view is important for allocating optimization efforts. In our case study, the API developers started by improving the performance of \epTranslationsColor and then later optimized the performance of \epFeedItemsColor. However, in order to understand the problems of the two endpoints, a big component represents understanding the outliers: because they are many. \va{the last sentence does not parse}

  \subsection{Automated Outlier Detection and Monitoring}
  \va{Shouldn't this be promoted to a section? It is neither utilization- nor performance- related. Either that, or Section~\ref{sec:user} should be folded into this section too, and then Sections~\ref{sec:version} and~\ref{sec:regression} merged into another section. Otherwise we have a very scattered paper}

  When an API is called from within a highly interactive application (as it is the case with the case study in this paper) 
  of particular interest to the API developers are performance {\em outliers}.   Indeed, a translation request that takes three times more than expected can seriously decrease the perceived quality of the application. Thus, identifying, collecting all appropriate data, and diagnosing the root causes of such outliers is especially critical in improving the quality of an application.

@@

  \begin{figure}[h!]
    \centering
    \includegraphics[width=0.9\linewidth]{outlier-annotated}
    \caption{Automatically collected outlier information}
    \label{fig:stack}
  \end{figure}

The bottom panel in \Fref{fig:stack} shows the stack trace. 
In this particular case, it is revealing for the developer to learn that at the time of the stack trace snapshot, the code was in the \code{google\_translator}: indeed, the system uses as back-end multiple translators, and it has been observed that many of the outliers happen while the system is waiting for the Google translator.
%
This information has to be corroborated with the observations that neither the memory nor the processor are overloaded at the moment. Thus this functionality in \code{microsoft\_translator} is really slow in itself, and this is not a result of the machine being overloaded for example. \va{last sentence does not parse}
